---
title: "Project CS544O2 - Analyzing NBA Myths"
author: "Samantha Kuang"
date: "12/4/2021"
output:
  html_document:
    fig_width: 8
    code_folding: hide
    highlight: tango
    toc: yes
    toc_float:
      collapsed: yes
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Objective - Examine Validity of Some "NBA Myths":

Many factors affect an NBA game. To understand the impact of each factor, three different data sets will be utilized. This enables us to analyze each games at player level as well as team level and check the rationality of certain "NBA Myths". The following questions will guide our analysis. 

### Questions: 

1. Shooting 
    a) Are players more likely to make shots from a particular location e.g. corner three.   
    b) <b>Hot Hand Analysis</b> - the belief that a player can get "hot" during a game and start making every 
    c) Warriors are known as "elite" shooters. How much farther do they actually shoot from?
    
2. <b>Home Court Advantage</b> - the belief that teams perform better at home then  away games. 
    a) Is there a location that's more or less challenging for teams to play at? E.g. Some believe that high elevation in Arizona gives the Nugget an advantage

3. Regular season do not matter - the belief that elite players do not play at full effort until the playoff
    a) What's the Warrior's likelihood of winning when Steph (their top player) is "locked in".


### Data Sets and Sources Information:
Conclusion from analysis is only a good as the data available. This project assumes that the data sets I am using are accurate reflections of the true historical NBA data.

1. <b> shots </b> : used python to extracted from <b> sports_radar </b> (unknown reliability)
    * Get nba schedule for the Warrior's last for games: '2021-11-30', '2021-12-03', '2021-12-04', '2021-12-06'
    * Provided game_id in the <b> playperplay database </b>. 
    * information about every shot attempted in all the games played in those 4 dates. 

2. <b> allnbagames </b> : used python to extract from the <b> rapidapi website </b>. (unknown reliability)
    * It has information about every nba game played between seasonyear 2017 to 2020 inclusive. 
    * "https://api-nba-v1.p.rapidapi.com/games/seasonYear/
    
3. <b> stephcurry_df </b> : from https://data.world/datatouille/stephen-curry-stats  (unknown reliability)
    * Has Steph Cury's box score information for eevery game between 2009 - 2017
    * "http://api.sportradar.us/nba/trial/v7/en/games/{}/pbp.json

```{r message=FALSE, warning=FALSE, include=FALSE}
# load/install required packages
library("httr")
library("readxl")
library(lubridate)
library(dplyr)
library(plotly)
library(tidyverse)
library(stringr)
```

# Hot Hand Analysis

### Data Ingestion and Wrangling
```{r echo=FALSE}
shots = read.csv("nba_shots.csv")

#Adding Columns for Enhancements
point_diff = c()
made_or_miss = c()

for (i in 1:nrow(shots)) {
  if (shots$EventType[i] == "twopointmade"){
    point_diff = c(point_diff, 2)
    made_or_miss = c(made_or_miss, "made")
  }else if(shots$EventType[i] == "threepointmade" ) {
    point_diff = c(point_diff, 3)
    made_or_miss = c(made_or_miss, "made")
  } else{
    point_diff = c(point_diff, 0)
    made_or_miss = c(made_or_miss, "miss")
  }
}
shots$point_contri = point_diff
shots$made_or_miss = made_or_miss


### Warriors DataSet
warriors = filter(shots, shots$Player %in% c("Chris Chiozza", "Jordan Poole", "Stephen Curry", "Draymond Green", "Andrew Wiggins", "Juan Toscano-Anderson", "Otto Porter Jr.", "Gary Payton II", "Damion Lee", "Kevon Looney", "Moses Moody", "Jonathan Kuminga", "Nemanja Bjelica", "Andre Iguodala")) 

#warriors %>% group_by(Player) %>% summarise(Num_Games = n_distinct(GameID), Num_Points = sum(point_contri), Median_Dist = median(Shot_Distance)) %>% arrange(desc(Num_Points))
```

In the data preparation step, added columns to the shots dataset for enhancement to enable ease of use for analyzing. Also created a subsetted of the shots data to analyze the Warriors more closely.


### All X and Y Coordinates of Shots Made
NBA court size is 94 X 50 feet. 3 point line is 23ft and 9 inches. Graphically visual of all points made given the timeline of our data set. 

```{r echo=FALSE}
threepoints = subset(shots, shots$EventType %in% 'threepointmade')
twopoints = subset(shots, shots$EventType %in% 'twopointmade')

fig1 <- plot_ly(threepoints, x = ~Locationx, y = ~Locationy, type = 'scatter', mode = 'markers', color = ~Player, title="layout.hovermode='closest' (the default)")
fig2 <- plot_ly(twopoints, x = ~Locationx, y = ~Locationy, type = 'scatter', mode = 'markers', title="layout.hovermode='closest' (the default)")
fig <- subplot(fig1, fig2, nrows = 2, margin = 0.07) %>% layout(title = "<b>NBA Games on 11-30, 12-03, 12-04, 12-06</b>")
annotations = list(
   list( 
    x = 0.05,  
    y = 1,  
    text = "Three Points Made",
    font=list(
                color="red",
                size=12
            ),
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),  
  list( 
    x = 0.05,  
    y = 0.45,  
    text = "Two Points Made",
        font=list(
                color="red",
                size=12
            ),
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ))
fig <- fig %>%layout(annotations = annotations) 
fig

```

### Spatial Correlation of Top Players
Any correlations between the x and y distance of their successful shoots?

1. "Giannis Antetokounmpo" 
2. "Kevin Durant"
3. "Chris Paul"  
4. "Jimmy Butler" 
5. "Trae Young"
6  "James Harden"
7. "LeBron James"
8  "Stephen Curry"

```{r echo=FALSE}
players = c("Giannis Antetokounmpo" , "Kevin Durant", "Chris Paul" ,"Jimmy Butler","Trae Young", "James Harden","LeBron James", "Stephen Curry")

spatial_corrs_made_shots = subset(shots, (shots$Player %in%  players) & shots$made_or_miss == "made")[,c("Player", "Locationx", "Locationy")] %>% group_by(Player) %>% summarise(corr_x_y_made = cor(Locationx, Locationy)) %>%arrange(Player)

spatial_corrs_miss_shots = subset(shots, (shots$Player %in%  players) & shots$made_or_miss == "miss")[,c("Player", "Locationx", "Locationy")] %>% group_by(Player) %>% summarise(corr_x_y_miss = cor(Locationx, Locationy))  %>% arrange(Player)


spatial_corrs_made_shots$corr_x_y_miss =spatial_corrs_miss_shots$corr_x_y_miss
spatial_corrs = spatial_corrs_made_shots

fig <- plot_ly(spatial_corrs, x=~Player, y=~corr_x_y_made, type = "bar", name = "Made Shots")
fig <- fig %>% add_trace(y = ~corr_x_y_miss, name = "Miss Shots")
fig <- fig %>% layout(title = "Correlation of X and Y Location of Top Player Shots", yaxis = list(title = "cor of x and y coordinates of shots"), barmode = "group")
fig
```
This shows where in the arc steph curry can make shots. 
1.Does this imply that Stephen Curry can only make shots at the top arc and not ?
2. Jimmy Butler makes a lot of his shots by the rim -so x and y are highly coorelated.


### Central Limit Theorem  - Mean Shooting Distance. 

```{r echo=FALSE}
set.seed(9566)
par(mfrow = c(2,2))
samples = 500
size = 10
size = 20
size = 30
size = 40

xbar_10 = numeric(samples)
xbar_20 = numeric(samples)
xbar_30 = numeric(samples)
xbar_40 = numeric(samples)

for (i in 1:samples){
  xbar_10[i] = mean(sample(shots$Shot_Distance, 10, replace = TRUE))
}
for (i in 1:samples){
  xbar_20[i] = mean(sample(shots$Shot_Distance, 20, replace = TRUE))
}
for (i in 1:samples){
  xbar_30[i] = mean(sample(shots$Shot_Distance, 30, replace = TRUE))
}
for (i in 1:samples){
  xbar_40[i] = mean(sample(shots$Shot_Distance, 40, replace = TRUE))
}

# graphs
vline <- function(x = 0, color = "grey") {
  list(
    type = "dash", 
    y0 = 0, 
    y1 = 1, 
    yref = "paper",
    x0 = x, 
    x1 = x, 
    line = list(color = color, alpha=0.6)
  )
}

fig1 <- plot_ly(x= xbar_10,type = "histogram", histnorm = "probability", alpha = 0.6, name = "Sample Size 10")%>% layout(shapes = list(vline(median(xbar_10))))

fig2 <- plot_ly(x= xbar_20,type = "histogram", histnorm = "probability", alpha = 0.6, name = "Sample Size 20")%>% layout(shapes = list(vline(median(xbar_20))))


fig3 <- plot_ly(x= xbar_30,type = "histogram", histnorm = "probability", alpha = 0.6, name = "Sample Size 30") %>%  layout(shapes = list(vline(median(xbar_30))))

fig4 <- plot_ly(x= xbar_40,type = "histogram", histnorm = "probability", alpha = 0.6, name = "Sample Size 40") %>%   layout(shapes = list(vline(median(xbar_40))))

fig <- plot_ly(x = ~rnorm(50),
             type = "histogram",
             histnorm = "probability")
fig <- subplot(fig1, fig2, fig3, fig4, nrows = 2, margin = 0.07) %>% layout(title = "Distance from the Hoop", 
            xaxis = list(range = c(0, 30)),
            xaxis2 = list(range = c(0,30)),
            xaxis3 = list(range = c(0,30)),
            xaxis4 = list(range = c(0,30)),
                  xaxis = list(title = "Shooting Distance"), yaxis = list(title = "Probability") )

fig

#comparing the sample mean and standard deviation
cat(str_c("The Mean Shooting Distance for the League is", round(mean(shots$Shot_Distance),4), "feet","\n" ,sep = " "))
cat(str_c(" sample size = 10", " Mean = ", round(mean(xbar_10), 4), "SD = ", round(sd(xbar_10),4), "\n", "sample size = 20", " Mean = ", round(mean(xbar_20), 4), "SD = ", round(sd(xbar_20),4), "\n",
"sample size = 30", " Mean = ", round(mean(xbar_30), 4), "SD = ", round(sd(xbar_30),4), "\n",
"sample size = 40", " Mean = ", round(mean(xbar_40), 4), "SD = ", round(sd(xbar_40),4), sep = " " )
)

```

  The Central Limit Theorem states that as we increase our sample size, our sample mean will reflect closer to the true population mean. Furthermore, the sample variance of our sample mean will decrease and the distribution will become normally distributed. As a result, we are more confident that our sample mean reflects the true population mean. 
  Graphically, we see this. The spread of our histogram decreases as we increase our sample size because of a smaller variance. Furthermore,the true mean is a shooting distance of 14.49feet, so the difference between the population mean and the sample mean becomes smaller as the sample size increases. 


### Examing Warriors Performance

Q) How much farther can the Warriors shoot compare to the league?

```{r echo=FALSE}
warriors %>% group_by(Player) %>% summarise(avg_dist = mean(Shot_Distance), 
                                            avg_points_per_attempt = sum(point_contri)/n(),
                                            total_points = sum(point_contri),
                                            num_games = n_distinct(GameID)) -> warriors_ppp
annotations = list(
   list( 
    x =48,  
    y = 25,  
    text = "Size of Dot represents Average Points Made per Attempt ",
    font=list(
                color="blue",
                size=12
            ),
     showarrow = FALSE ))

fig <- plot_ly(warriors_ppp, x = ~total_points, y = ~avg_dist, text =~warriors_ppp$Player)
fig <- fig %>% add_markers(size = ~avg_points_per_attempt)
fig <- fig %>% add_text(textposition = "top right")
fig <- fig %>% layout(title = "<b> Warriors' Players - Performance Statistics </b>",
                      xaxis = list(range = c(0, 100)),
                      annotations = annotations,
                     showlegend = FALSE)

fig
```
Conclusion:
1. Steph Curry generates the most points, followed by Andrew Wiggins and Jordan Poole. He has a lower average points made per attempt tahn other players, but this may be due to a multitude of reasons. 
- having more challenging looks (guarded shots)
- providing good assists to this teammates
- shooting farther


### Sampling 

There are many sampling methods and the sample results are used to estimate the population characteristics. Employing several sampling methods, the follow analysis is aim to understand the true shooting distance of the Warriors Team. Comparing these numbers with the Leagues' shooting average will help us quantify how strong a
shooting Team the Warriors is. 


```{r echo=FALSE}
library(sampling) 
set.seed(9566)

#not enough data point for these players. To prevent stratum sampling error: 
warriors = subset(warriors, !(warriors$Player %in% c("Moses Moody","Chris Chiozza") ))
warriors= drop_na(warriors)


#simple random sampling without replacement
x= srswor(100,nrow(shots))
srswor_x = shots[x != 0,]
freq_x = table(srswor_x$Action_Area)

#systematic sampling
N = nrow(warriors)
n = 100
k = floor(N/n)
r = sample(k, 1)
s= seq(r, by = k, length = n)
systematic = warriors[s,]

#inclusion probabilities  - systematic sampling with unequal probabilities
pik = inclusionprobabilities(warriors$Shot_Distance, 100) 
s = UPsystematic(pik)
sample_inclusion = warriors[s != 0, ]


#stratified sampling - ordered data
ordered_shots = warriors[order(warriors$Player),] 
sample_size = 100
freq = table(ordered_shots$Player)
#doing some rounding here to get sample size = 50
strata_size = ceiling(sample_size * freq/sum(freq))
sample_strata = strata(warriors, 
                       stratanames = c("Player"),
                       size = strata_size, 
                       method = "srswr")
sample_strata = getdata(warriors, sample_strata) 

#Graph
fig = plot_ly(y = as.vector(warriors$Shot_Distance), type = "box", name = "Warriors Data", title="layout.hovermode='closest' (the default)") %>%  
  add_trace(y = as.vector(sample_strata$Shot_Distance), type = "box", name="Stratified Sampling") %>%
      add_trace(y = as.vector(sample_inclusion$Shot_Distance), name = "UPsystematic Method") %>%
      add_trace(y = as.vector(systematic$Shot_Distance), name = "Systematic Sampling") %>%
      add_trace(y = as.vector(srswor_x$Shot_Distance), name = "SRSWOR") %>% layout(title = "Warriors Shooting Average - Sampled on Players")
      
fig

#Details
cat(str_c("The Mean Distance for the GS Warriors is ", round(mean(warriors$Shot_Distance),2)," feet "
,"\n", 
"Mean from Stratified Sampling of the CSG is ", round(mean(sample_strata$Shot_Distance), 2), "\n",
          "Mean from Sample Inclusion is ",  round(mean(sample_inclusion$Shot_Distance,2), 2), "\n",
          "Mean from Systematic Sampling is ", round(mean(systematic$Shot_Distance), 2), "\n",
"Mean from SRSWOR is ", round(mean(srswor_x$Shot_Distance),2), sep = " " ))

```

Conclusion: 

1. It looks like the most accurate is Systematic Sampling, follow by SRSWOR, then Stratified Sampling then the UPsystematic method. This could be due to the fact that the data is ordered by games. Players perform differently in each game - taking more or less shots /shooting better or worst. So Systematic Sampling may have drawn an equal number of shots from each game. 


### Hot Hand - Wait Time until next made shot after shot has been made - Exponential Distribution
Is the <b> Wait Time </b> normally distributed? This analysis of hot hand would imply that there's a peak closer to 0. 

```{r echo=FALSE}
# "Nikola Jokic"
Jokic = filter(shots, Player == "Nikola Jokic" & GameID == "b8ebbc5a-75cf-4b4f-b876-5a907e315d8c" )
Jokic_wait_time = c()
counter = 0
for (i in 1:nrow(Jokic)) {
  if (Jokic$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    Jokic_wait_time = c(Jokic_wait_time,counter)
    counter = 0 #resetting counter to 0
  }
}

#BootStrapping Method
resample_1 = replicate(1000, sample(Jokic_wait_time, replace = TRUE), simplify = FALSE)
sample_mean_star_jokic = sapply(resample_1, mean, simplify = TRUE)

# Kevin Durant
Durant_g1 = filter(shots, Player == "Kevin Durant" & GameID == "a77dab55-c46b-438a-af52-98eb7aa121d2" )
Durant_wait_time = c()
counter = 0
for (i in 1:nrow(Durant_g1)) {
  if (Durant_g1$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    Durant_wait_time = c(Durant_wait_time,counter)
    counter = 0 #resetting counter to 0
  }
}
Durant_g2 = filter(shots, Player == "Kevin Durant" & GameID == "26fec744-173b-42ac-9e21-8e07ace0d83b" )
counter = 0
for (i in 1:nrow(Durant_g2)) {
  if (Durant_g2$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    Durant_wait_time = c(Durant_wait_time,counter)
    counter = 0 #resetting counter to 0
  }
}
Durant_g3 = filter(shots, Player == "Kevin Durant" & GameID == "a158afce-9043-4c03-bd3f-eb64fc6759fe" )
counter = 0
for (i in 1:nrow(Durant_g3)) {
  if (Durant_g3$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    Durant_wait_time = c(Durant_wait_time,counter)
    counter = 0 #resetting counter to 0
  }
}

#BootStrapping Method
resample_1 = replicate(1000, sample(Durant_wait_time, replace = TRUE), simplify = FALSE)
sample_mean_star_durant = sapply(resample_1, mean, simplify = TRUE)

# "Stephen Curry"
curry_g1 = filter(shots, Player == "Stephen Curry" & GameID == "8daf430b-7e49-4b8c-a176-74c19f2c8f6b" )
curry_g2 = filter(shots, Player == "Stephen Curry" & GameID == "981f30b5-11a8-4611-a87f-4a08b33f762f" )
curry_g3 = filter(shots, Player == "Stephen Curry" & GameID == "78873b07-9a84-4d23-8e27-0aa58cd09cb1" )
curry_g4 = filter(shots, Player == "Stephen Curry" & GameID == "d16a0731-3c62-4369-a223-4d313bac4e83" )

curry_wait_time1 = c()
counter = 0
for (i in 1:nrow(curry_g1)) {
  if (curry_g1$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    curry_wait_time1 = c(curry_wait_time1,counter)
    counter = 0 #resetting counter to 0
  }
}

curry_wait_time2 = c()

counter = 0
for (i in 1:nrow(curry_g2)) {
  if (curry_g2$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    curry_wait_time2 = c(curry_wait_time2,counter)
    counter = 0 #resetting counter to 0
  }
}

curry_wait_time3 = c()
counter = 0
for (i in 1:nrow(curry_g3)) {
  if (curry_g3$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    curry_wait_time3 = c(curry_wait_time3,counter)
    counter = 0 #resetting counter to 0
  }
}

curry_wait_time4 = c()
counter = 0
for (i in 1:nrow(curry_g4)) {
  if (curry_g4$made_or_miss[i] == "miss"){
    counter = 1+ counter
  } else {
    curry_wait_time4 = c(curry_wait_time4,counter)
    counter = 0 #resetting counter to 0
  }
}

curry_wait_time = c(curry_wait_time1, curry_wait_time2, curry_wait_time3, curry_wait_time4)

sample_mean_curry_wait_time = mean(curry_wait_time)


#BootStrapping Method
resample_2 = replicate(1000, sample(curry_wait_time, replace = TRUE), simplify = FALSE)

sample_mean_star_curry = sapply(resample_2, mean, simplify = TRUE)
#hist(sample_mean_star, breaks = 50)


#Graphing
fig <- plot_ly(alpha = 0.7) %>% 
  add_histogram(x = ~sample_mean_star_jokic,
                name = "Nikola Jokic") %>% 
  add_histogram(x = ~sample_mean_star_curry,
                name = "Stephen Curry") %>%
  add_histogram(x = ~sample_mean_star_durant,
                name = "Kevin Durant") %>%
  layout(barmode = "overlay",
         title = "Wait Time Until Made Shots",
         xaxis = list(title = "Number of Shots before Made",
                      zeroline = FALSE),
         yaxis = list(title = "Count",
                      zeroline = FALSE))
fig

# stats
sample_mean_jokic = mean(Jokic_wait_time)
sample_mean_durant = mean(Durant_wait_time)
cat(str_c("Mean Wait Time for Curry = ", round(sample_mean_curry_wait_time,2), " Number of Shots", "\n",
"Mean Wait Time for Durant = ", round(sample_mean_durant,2), " Number of Shots", "\n",
"Mean Wait Time for Jokic = ", round(sample_mean_jokic,2), " Number of Shots", "\n"))
```
```{r echo=FALSE}

x = subset(shots, (shots$Player %in% c("Stephen Curry", "Kevin Durant")) & (shots$EventType == "threepointmade"))%>% group_by(Player)%>% summarise(avg_three_per_game= n()) 


y = subset(shots, (shots$Player %in% c("Stephen Curry", "Kevin Durant")) & (shots$EventType == "threepointmade"))%>% group_by(Player, GameID)%>% summarise(avg_three_per_game= n()) 

x$avg_three_per_game/table(y$Player)

```

Conclusions:
Durant and Jokic has a shorter wait time until their next made shot compare to Curry. However Curry takes more three points. This would mean that every successful shot of Durants and Jokic amounts to 2 points while Curry's made shots can contribute 2 - 3 points. 

It looks like Jokic has a short wait time until his next successful shot given that he has made a shot. However, this was based on 1 game for Jokic, explaining the white spaces between the blue bars. So, the hot hand myth is inconclusive given the lack of data. 

Stephen Curry averages 5.25 threes per game while Kevin Durant only averages 2 threes per game. 


# Analyzing Home Court Advantage 

### Data Ingestion, Explore Data, Perform preprocessing and Data Wrangling
```{r echo=FALSE}
par(mfrow = c(3,1))

#dataset1 = allnbagames to answer homecourt advanatage
nbagames = read.csv('nba_games_exportedon_for_rproject.csv')

#Length of vectors, class, and mode of each column
par(mfrow = c(3,1))
barplot(table(nbagames$seasonYear), main = "Number of Games Played Per Season", col = c("cyan"))
barplot(table(nbagames$hTeam.nickName), main = "Number of Home Games Played From 2015 - 2020", col = c("red"), las = 2)
barplot(table(nbagames$vTeam.nickName), main = "Number of Away Games Played Per 2015 - 2020", col = c("blue"), las = 2)
```

After examining the dataset, there are irrelevant records in the dataset

1. row filter- only keeping:
- league = standard
- country = US, USA, and Canada(Toronto Raptors)
- gameStatus = Finish

2. Some attributes needs to be converted to the proper data type. The following are Identified:
- startTimeUTC and endTimeUTC needs to be converted to datetime 
    - redefining column gameduration to endTimeUTC - startTimeUTC in minutes
    - current gameduration is unusable because it is a string in hours and minutes
- Calculating a new column call GameDate by converting startTimeUTC to Date 

3.  Need to filter out columns that are irrelevant or has replicated information

4. not complete data for 2015 or 2016. So we will be focusing on 2017 only.

```{r message=FALSE, warning=FALSE, include=FALSE}
#Filtering out irrelevant data
nba = subset(nbagames, (nbagames$league %in% "standard") & (nbagames$country %in% c("USA", "Canada", "US")) & (nbagames$statusGame %in% "Finished"))
#table(nba$country, nba$league, nba$statusGame)

#Converting gameduration from string to minutes  
nba$startTimeUTC= as.POSIXct(nba$startTimeUTC,format="%Y-%m-%dT%H:%M:%OS")
nba$endTimeUTC= as.POSIXct(nba$endTimeUTC,format="%Y-%m-%dT%H:%M:%OS")
nba$gameDuration = difftime(nba$endTimeUTC, nba$startTimeUTC, units = "mins")

#Creating Variable "GameDate"
nba$GameDate= as.Date(nba$startTimeUTC)

#Examing nbagames from year 2017 - 2018 only because of non complete data
nba = subset(nba, nba$GameDate >  '2017-09-01')
nba = subset(nba, nba$GameDate <  '2018-09-01')

#removing noise teams
nba = subset(nba, !(nba$vTeam.nickName %in% c("Sharks","Maccabi Haifa","Long-Lions","United", "Bullets")))

#Only keeping relevant columns and relevant teams
nba = nba[, c("seasonYear", "league","arena", "city", "country",
              "GameDate","gameDuration", "vTeam.nickName","vTeam.score.points",
              "hTeam.nickName", "hTeam.score.points")]

#Removing teams that are  not pat of NBA
nba = subset(nba, !(nba$hTeam.nickName %in% c("Team Durant", "USA", "")))



WinningTeam = c()
LossingTeam = c()

for (i in 1:nrow(nba)){
  if (nba$hTeam.score.points[i] > nba$vTeam.score.points[i]){
    WinningTeam = c(WinningTeam, nba$hTeam.nickName[i])
    LossingTeam = c(LossingTeam, nba$vTeam.nickName[i])
  }
  else{
    WinningTeam = c(WinningTeam, nba$vTeam.nickName[i])
    LossingTeam = c(LossingTeam, nba$hTeam.nickName[i])

  }
}

nba$teamwon = WinningTeam
nba$teamloss = LossingTeam

```

```{r message=FALSE, warning=FALSE, include=FALSE}
homewon = nba %>% group_by(hTeam.nickName, teamwon) %>% filter(hTeam.nickName == teamwon ) %>% summarise(count = n()) %>% arrange(hTeam.nickName)

homeplayed = nba %>%group_by(hTeam.nickName) %>%summarise(count = n())  %>% arrange(hTeam.nickName)

home_stats = data.frame(Team = homewon$teamwon, Wins_Home = homewon$count, Home_Games = homeplayed$count)

awaywon = nba %>%group_by(vTeam.nickName, teamwon) %>% filter(vTeam.nickName == teamwon ) %>%summarise(count = n())
awayplayed = nba %>%group_by(vTeam.nickName) %>%summarise(count = n())  %>% arrange(vTeam.nickName)

away_stats = data.frame(Team = awaywon$teamwon, Wins_Away = awaywon$count, Away_Games = awayplayed$count)
 


merge_df = merge(x = home_stats, y = away_stats, all.x= TRUE)

merge_df$Losses_Home = merge_df$Home_Games - merge_df$Wins_Home
merge_df$HomeWinningPercentage = merge_df$Wins_Home/ merge_df$Home_Games

merge_df$Losses_Away = merge_df$Away_Games - merge_df$Wins_Away
merge_df$AwayWinningPercentage = merge_df$Wins_Away/ merge_df$Away_Games


merge_df$True_Winning_Percentage = (merge_df$Wins_Home + merge_df$Wins_Away)/(merge_df$Home_Games+ merge_df$Away_Games)
```



### For each team -  what's the percentage of winning at home home games?

```{r}
fig <- plot_ly(merge_df, x = ~Wins_Home, y = ~Team, type = 'bar', orientation = 'h', name = 'HomeGameWin',
        marker = list(color = 'rgba(246, 78, 139, 0.6)',
                      line = list(color = 'rgba(246, 78, 139, 1.0)',
                                  width = 1)))
fig <- fig %>% add_trace(x = ~Losses_Home, name = 'HomeGameLoss',text = paste0(round(merge_df$HomeWinningPercentage,2)*100,"%"), textposition = 'outside',
            marker = list(color = 'rgba(58, 71, 80, 0.6)',
                          line = list(color = 'rgba(58, 71, 80, 1.0)',
                                      width = 1)))
fig <- fig %>% layout(barmode = 'stack',
        title = "Number of Wins and Loss at Home Games for all NBA season 2017 - 2018",
         xaxis = list(title = "Number of Home Games"),
         yaxis = list(title ="NBA Teams")
        )
fig

```



```{r}
cat(paste("On average, NBA teams has a ", round(mean(merge_df$HomeWinningPercentage) * 100,0),"%", " chance of winning at Home Game, giving them a slight advantage over their opponents.", sep =""))
```

This is only one perspective. There may be many confounding variables e.g. good teams will win at home and away - masking the advantage or home court advantage if there is one. So we cannot make a conclusive statement.


### Does the location affect outcomes?

```{r}
location_df = nba
row.names(location_df) <- 1:nrow(location_df)
#ploting location of 
library(maps)
data(us.cities)

remove_cities = c('Cleveland OH', 'North Atlanta GA', 'West New York NY', 'North Miami Beach FL', 'North Miami FL', 'Portland ME', 'Port Charlotte FL', 'North Las Vegas NV', 'Kansas City KS', 'Seattle Hill-Silver Firs WA', 'South San Francisco CA', 'West Sacramento CA', 'East Los Angeles CA', 'Miami Beach FL')

us.cities = subset(us.cities, !us.cities$name %in% remove_cities)

location_df = nba[, c('seasonYear','hTeam.nickName','vTeam.nickName', 'arena', 'city')]

idx2 <- sapply(location_df$city, grep, us.cities$name)
idx1 <- sapply(seq_along(idx2), function(i) rep(i, length(idx2[[i]])))

location_df = cbind(location_df[unlist(idx1),,drop=F], us.cities[unlist(idx2),,drop=F])

#graphical analysis
graph_loc = unique(location_df[c("city", "name","arena", "lat", "long")])

arena_wins = nba %>% group_by(arena, teamwon, hTeam.nickName
) %>% filter(teamwon == hTeam.nickName) %>% summarise(Wins_in_Arena= n())

geo_wins = merge(x = graph_loc, y = arena_wins, by= "arena", all.x = TRUE)

#GRAPHING WINS BY AREA ON MAP
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray95"),
  subunitwidth = 1,
  countrywidth = 1,
  subunitcolor = toRGB("white"),
  countrycolor = toRGB("white")
)

fig <- plot_geo(geo_wins, lat = ~lat, lon = ~long)

fig <- fig %>% add_markers(
    text = ~paste(arena, city,sep = "<br />"),
    size = ~Wins_in_Arena, hoverinfo = "text"
  )
fig <- fig %>% colorbar(title = "Games")
fig <- fig %>% layout(
    title = 'US Map of NBA locations', geo = g
  )
fig

```

To eliminate others factors that may be affecting the impact of player's performance in different locations. Denver is the primarly focus. The objective is to calculate the difference between chance of winning for every team not at Devener vs at Denver. 


```{r}
# what's this team's winning percentage at denver? 
denver = subset(nba, arena %in% "Pepsi Center")
teamwon = denver %>% group_by(teamwon) %>% summarise(count = n())
teamvisitng = denver %>% group_by(vTeam.nickName) %>% summarise(count = n())
colnames(teamvisitng) = c("team", "num_games")
colnames(teamwon) = c("team", "num_games_won")

winningatdenver = merge(x = teamvisitng, y = teamwon, by="team" , all.x = TRUE) 

winningatdenver[is.na(winningatdenver)] = 0

winningatdenver$winning_percentage = winningatdenver$num_games_won/winningatdenver$num_games

cat(str_c("League's likelihood of winning at Denver is ", round(mean(winningatdenver$winning_percentage),4)*100, "%", "\n", "League's likelihood of winning at Away Game is ", round(mean(merge_df$AwayWinningPercentage),4)*100,"%", "\n",
"SD is ", round(sd(merge_df$AwayWinningPercentage),4)*100,"%"))

```
Even though teams are less likely to win at Denver, 30.46% is within 1 standard derivation of 40.9% (+/- 14.5%). This means we cannot make any conclusions with high confidence.


```{r}
fig = plot_ly(type = 'box')
fig = fig %>% add_boxplot(y = merge_df$HomeWinningPercentage, quartilemethod="linear", name="% of Winning at Home Games",
                          jitter = 0.3, pointpos = -1.8, boxpoints = 'all')
fig = fig %>% add_boxplot(y = merge_df$AwayWinningPercentage, quartilemethod="linear", name="% of Winning at  Won at Away Games",
                           jitter = 0.3, pointpos = -1.8, boxpoints = 'all')
fig =  fig %>% layout(title = list(text = "Distribution of Winning Percentage for Teams"),
                      yaxis = list(range=c(0,1)
                                )
                      )
fig
```

This supports the idea that there's high variance in Winning Expectation for Teams. 


### Examing just the Warriors to see if there is a drop in performance given travel and timezone difference. 

```{r}
#warriors performance at home came ver away game:
warriors = nba[ (nba$hTeam.nickName == "Warriors" ) | (nba$vTeam.nickName == "Warriors"), ]

#warriors performance at home games:
warriorshome <- warriors[which(warriors$hTeam.nickName == "Warriors"),]
homedensity <- density(warriorshome$hTeam.score.points)

#warriors performance at away games:
homedensityaway = warriors[which(warriors$vTeam.nickName == "Warriors"), ]
awaydensity = density(homedensityaway$vTeam.score.points)

vline <- function(x = 0, color = "grey") {
  list(
    type = "dash", 
    y0 = 0, 
    y1 = 1, 
    yref = "paper",
    x0 = x, 
    x1 = x, 
    line = list(color = color, alpha=0.6)
  )
}

fig <- plot_ly(x = ~homedensity$x, y = ~homedensity$y, type = 'scatter', mode = 'lines', name = 'Points Made at Home Games', fill = 'tozeroy') %>% layout(shapes = list(vline(median(homedensity$x)), vline(median(awaydensity$x))))

fig <- fig %>% add_trace(x = ~awaydensity$x, y = ~awaydensity$y, name = 'Points Made at Away Games', fill = 'tozeroy')
fig <- fig %>% layout(xaxis = list(title = 'Density of the Points Distribution'),
         yaxis = list(title = 'Density'), 
         title = list(text = "Warriors: Points Distribution at Home vs Away Games between 2015 - 2020", y = 0.95))
fig

cat(str_c("The median of the GSW Total Score is " , median(awaydensity$x), ". \n", "The median of the GSW Total Score at Home Games is", median(homedensity$x), ". \n"))
```
Warrior's scoring abilities doesn't seem to be affected by away games significantly. Only a 4 points difference between at home vs away.


### Data Ingestion and Preprocessing for Stephen Curry's Game stats 

This dataset will be used to compare Steph Curry's player performance during regular season games vs playoff games. For accuracy of performance, the data will be filtered on seasons that the Warriors made it to the playoffs. 

```{r message=FALSE, warning=FALSE, include=FALSE}
GET("https://query.data.world/s/g373rfhyxepvm42ew3763gpbz6wtwu", write_disk(tf  <- tempfile(fileext = ".xlsx")))
curry_df <- read_excel(tf)
head(curry_df, n = 5)


#Data Cleaning
curry_df = na.omit(curry_df)
curry_df = curry_df[curry_df$Minutes != 0, ]
```

```{r message=FALSE, warning=FALSE, include=FALSE}
curry_df$yearmonth = format(curry_df$Dates, format = "%Y-%m")
curry_df$year = format(curry_df$Dates, format = "%Y")

table(curry_df$Type, curry_df$year)
noise = c("2018-10","2012-01", "2012-02", "2012-03", "2012-04")
made_year = 2012:2018

# subsetting data on 2013 - 2018
seasons_in_playoff = subset(curry_df, (curry_df$year %in% made_year) & !(curry_df$yearmonth %in% noise) & (!is.na(curry_df$Type)) & (curry_df$Minutes != 0) )
seasons_in_playoff
```


# Performance During Regular vs Playoff Games
- Is there a performance difference between Regular Season Games and Playoff Games for Top Players?
- Focusing only on Steph Curry for this analysis.

```{r}
# What is Curry's shooting percentage during regular season - including fg and ft
regular = seasons_in_playoff[which(seasons_in_playoff$Type == "REGULAR SEASON STATS") ,]
shooting_percentage = regular$`Successful Shots`/regular$`Total Shots`
dates = regular$Dates
regdf = data.frame(dates, shooting_percentage)

# What is Curry's shooting percentage during conference and finals
postseason = subset(seasons_in_playoff, !(seasons_in_playoff$Type %in% "REGULAR SEASON STATS"))
shooting_percentage = postseason$`Successful Shots`/postseason$`Total Shots`
dates = postseason$Dates
psdf = data.frame(dates, shooting_percentage)

##

#summary(regdf$shooting_percentage)
#summary(psdf$shooting_percentage)
cat(paste("Regular seasson shooting pecentage:", sd(regdf$shooting_percentage), ". \n",
      "Post reson shooting percentage: ", sd(psdf$shooting_percentage), ". \n", 
"since the sd is relatively the same, a t-test is applied. "))
# T- test psdf and regdf
t.test(regdf$shooting_percentage, psdf$shooting_percentage)
cat(paste("H0: µ1 = µ2 (psdf and regdf means are equal)", "\n","HA: µ1 ≠ µ2 (psdf and regdf  means are not equal)"))
```
Conclusion: 
The p-value = 0.08 from the Welch Two Sample t-test. There's no significant shooting difference between regular season games and playoff games for Steph Curry. 

```{r}
fig <- plot_ly(x=~regdf$dates, y=~regdf$shooting_percentage, type = 'scatter', name = "regular season games")
fig <- fig %>% add_trace(x=~psdf$dates, y =~psdf$shooting_percentage, name = "playoff games") %>%layout(title = "Shoot Percentage by Dates", yaxis = list(title ="Shooting Percentage"), xaxis = list(title ="Game Dates"))
fig
```
The spread of Steph Curry's shooting percentage is relatively similar between Regular Season Games and Playoff Games. However, there are more variance in the Regular Season Games.

### Examining Corrlation between Steph Curry PTS Contributions and Team Score:
```{r}
regular$`Score GS` = as.numeric(regular$`Score GS`)
r = regular$Result
x = c()
for (i in 1:length(r)) {
  if (r[i] == "W"){
    x= c(x,1)
  } else{
    x = c(x, 0)
  }
}  

stats = regular[,c("Score GS",  "3 Points Succesful","PTS", "REB", "AST", "BLK", "STL","TO",  "Minutes")]
stats$Result_Encoded = x
regcorr = cor(stats)

#postseason
postseason$`Score GS` = as.numeric(postseason$`Score GS`)
r = postseason$Result
x = c()
for (i in 1:length(r)) {
  if (r[i] == "W"){
    x= c(x,1)
  } else{
    x = c(x, 0)
  }
}  

psstats = postseason[,c("Score GS",  "3 Points Succesful","PTS", "REB", "AST", "BLK", "STL","TO",  "Minutes")]
psstats$Result_Encoded = x
pscorr =cor(psstats)

fig1 <- plot_ly(x=colnames(regcorr), y=rownames(regcorr), z = regcorr, type = "heatmap", color= c("cyan", "blue")) %>%
    layout(margin = list(l=120),  color= c("cyan", "blue"))


fig2 <- plot_ly(x=colnames(pscorr), y=rownames(pscorr), z = pscorr, type = "heatmap", color= c("cyan", "blue")) %>%
    layout(margin = list(l=120))


fig <- subplot(fig1, fig2, nrows = 2, margin = 0.07) %>% layout(title = "Correlation HeatMap Reg vs Post")
fig
```

Result_Encoded = 1 if Win, 0 if Loss. 
The correlation heatmap did not reveal any significant correlation between winning and another factor. Take aways is that Wins are positively correlated with Warriors' Team Scores. 

Small correlations to Point Out:
1. Positively correlated with Steph Curry's points and Rebounds.
2. Negatively correlated with Steph Curry's Turn Overs and Playtime. 
  * Further analysis on this topic is to check if Steph is more likely to make more mistakes as he plays longer due to exhaustion and lost of focus. 

### Impact of Not Resting Top Players

```{r}
xreg = seq(15, 60, by = 5)
p_win_reg = c()
  
for(i in xreg){
  testdf = subset(stats, (stats$Minutes >= i) & (stats$Minutes < i+5))
  probwins = sum(testdf$Result_Encoded)/nrow(testdf)
  p_win_reg = c(p_win_reg, probwins)
}


xps = seq(15, 60, by = 5)
p_win = c()
  
for(i in xps){
  testdf = subset(psstats, (psstats$Minutes >= i) & (psstats$Minutes < i+5))
  probwins = sum(testdf$Result_Encoded)/nrow(testdf)
  p_win = c(p_win, probwins)
}

fig = plot_ly(x = xps, y = p_win, type = "scatter", name = "playoffs")
fig = fig %>% add_trace(x=xreg, y= p_win_reg, name = "regular seasons") %>% layout(title = "Chance of Win vs Total Game Time", yaxis = list(title = "Probability of Winning"), xaxis = list(title = "Total Game TIme"))
fig

```

There appears to be a linear negative relationships between Steph Curry's total time played that the Warriors' chance of winning. This can be caused by the depth of the Warriors' Rosters. When the team has no depth, and players are injured, Steph Curry has to play more minutes which decreasing their chance of winning.


# Project Conclusion:

May factors affects a team's performance. In examining the many questions above, we attempt to verify the legitimacy of many NBA Myths. 

1. Hot Hand doesn't appear to apply for Steph Curry or Kevin Durant. They appear to to be consistent players who makes a successful shot on an averaged of 1-2 attempts. For Jokic, there was not enough data to conclusively state that he is a streaky player. He could just be taking easy shots and making smarter decisions. The x and y coordinates of Jimmy Butler's shots appears to be highly correlated. A plausible explanation of this is that he's a rim put back player or a dunker, not a shoot. So he's really close to the basket. 

2. The Home Team does appear to have a slight advantage. Furthermore, teams appear to have a challenging time winning at Denver in the 2017 season. However, since there's very high variance in the mean associated with home court advantage, we cannot conclude that it's valid. Many confounding variabels are at play, e.g. some teams are more dominant given their makeup of players. 

4. Steph Curry is known as an elite player. In the 5 years that he made the playoffs, there is no significant difference between his performance during the regular vs post season. (Granted that he won MVP for two of those seasons, proving that he was playing well during the regular seasons.) There appears to be a  negative correlation given the Warriors' chance of winning and Steph Curry's play time. Further analysis needs to be conducted to make any conclusive statemetns. Could this be attributed to player exhaustion and loss of focus during long games? Maybe there's no depth in the team roster and other star players are injured/out? 



